### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
terminada revisao do manuscritp de tp

\section{Conclusions and further work}\label{conc}
This document presented a potentially novel description of Toki Pona (TP)
language in Section~\ref{basics}, and innovative and useful software routines for dealing with
the language in Section~\ref{hacks}.
As a minimalist conlang, TP is very convenient for cognitive
experiments
through linguistic semantics and
for devising tools for analysis, creation and visualization.
Other conlangs, and even non-planned languages might benefit from
TP and the content presented here in many ways, e.g. one might
synthesize TP texts and translate them as convenient,
adapt the routines to obtain statistics of the vocabulary,
or take advantage of the language description to devise new conlangs
or stilistic guidelines for any language.

Possible next steps are numerous:

@@

  \item Further develop the text synthesis facilities described in

@@

  \item Make a syntax highlighting that colors the tokens with respect
    to the syntatic position and function, and not in a fixed manner
    as is implemented and exposed in Section~\ref{shigh}.
    This might be implemented by using n-grams and
    further techniques from Natural Language Processing
  \item Enhance and study the TP Wordnet described in Section~\ref{wn}:
    implement a NLTK-like Wordnet interface; check if each synset

@@

    semicolon; find out the neighborhood of the synsets of the TP
    Wordnet in the English Wordnet; explore the possibilities for
    devising new conlangs by means of criteria related to Wordnet.

@@

    use insights derived from TP.
    Maybe create conlangs for specific uses:

@@

    A conlang might have different modes that might 
    be specified by a section header or tag. 
  \item Know about previously existing words that were used for TP

@@

    and about the reasons that led Sonja (and maybe other people)

@@

  \item Make an article written in TP

@@

    facilitating the acquisition of context,
    and a article in a canonical structure
    (introduction and related work, materials and methods,

@@

    I first conceived something around complexity, statistics, physics, or computer science.
    But one possibility is to write about linguistics, philosophy, literature
    or psychology with the partners I write in English.
    I can start a draft,
    they might learn the language in a few hours or weeks
    (with or without my help),
    to contribute, and we can write a short paper.
  \item Enhance the synthesis of text to yield better contextualized text
    and stylistic traces.
    Give the user the ability to choose the sentences:
    generate a sentence randomly according to
    previously written or given text,
    to some rules input by the user,
    to the package and the language guidelines.
    Should output to the screen and asks to keep and discard.

@@

  \item We are particularly interested in software tools,
    and one possibility is to use a conlang to describe software

@@

    In this sense, it seems profitable

@@

    Such study will probably entail considerations about
    parsing TP.

@@

\appendix
\section{My usage of Toki Pona}\label{mytoki}
I use the standard sounds, but often use [z] for s.
I often translate texts to Toki Pona (e.g. biblical excepts)
and create new texts as poems and short stories.
Most of them are in~\cite{tokisona}
and/or published in the TP Facebook groups.
I omit the li particle after subjects sina and mi,
in accordance with the norm,
but sometimes I use them when there are many predicates.
E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
In such cases, the first li is sometimes omitted.
Also, sometimes I use li after mi and sina where I find
that there is unwanted ambiguity, e.g.
sina moku pona 
which 
might be 'sina li moku pona' or 'sina moku pona (li jo e sike)'.

Names are by default transliterated,
but I advocate that, as in other languages,
names might be used as they are in the
correspondent mother tongue.
E.g. the name Erdös is used in
English and Portuguese although the standard
alphabet does not contain ö in such languages.
I also tend to legitimate the use of English (or German) words
in TP texts if it is the case,
as happens often in scientific writing
(kernel is a German word used in English,
webpage and software are English words used in Portuguese).

Proposed notations for numbers seem numerous.
I tend to think that one might indicate if two numbers
are being multiplied (pi) or are in different scales
(such as in decimal or binary notations).
For example, luka (pi) two might mean 52.
Most importantly,
52 is a reasonable notation for a simple language.
E.g.
'mi jo e jan sama nanpa 12',
or
'ona li lon e soweli 27'.

Avoiding needless words...
I've been avoiding e ni: and using only :.
I've been omitting the subject (sometimes also the li)
if it is the
same as in the last sentence.
I tend to group these words:
noka and anpa; luka, suno, sike and lawa; pali and pana.
These clusters of meaning hint me that an even simpler
language is possible and maybe more optimized to the
core goals of TP.

Also, ambiguities introduced by omission of li
and absence of a token to denote preposition,
suggest to me the possibility of a syntax that is always uniquely
parsed (or at least less ambiguous).

I have been inclined to conceive a language with the same syntax of TP:
'sentence' la 'sentence = subject + predicate + object (each term with
a possible preposition complement).
but always using particles to bound the sentence sections,
repeating them when the function is performed in the smallest scale:
'toki pi toki pona pi jan sona' meaning
language in toki pona and of intellectuals vs
'toki pi toki pona pi pi jan sona'
meaning language in [toki pona of intelectuals].
About the keyword for initiating a preposition complement 
might be 'a', but it is already taken in TP.
Maybe use 'a' before a preposition and use 'ha' or 'he'
instead of TP a.
In such a setting,
one might enable pi li e a (concept-qualifier, subject-predicate,
predicate-object, concept-preposition)
in any slot of the template:
'sentence' la 'sentence = subject + predicate + object (each term with
a possible preposition complement)'
Between sentences one would use:
la pi, la li, la e, la a, and la la.
This is a very fractal conlang proposal.
Maybe also have a way to discern between a noun,
a qualifier (adjectives/adverbs) and a verb,
and accept any of them for the subject, predicate, object, preposition,
slots. Or assume a part of speech as default for a slot or for the
first word in a slot.

\section{Final words in Toki Pona}\label{ftp}
toki li nasin e lawa.
li nasin tawa (pi) toki insa.

% jan li ken toki li ken pali pona e toki pona.
toki pona li pona e nasin tan ni:
ona li jo e nimi nanpa lili.
ona li pona.
pona kepeken weka 'p' li ona, a.

o taso la toki pona li kalama li lukin pona.
sitelen en nasin li open e sitelen
suli~\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
li sona. li nasin e toki e sona e lawa e lon.

toki ni li wawa tawa jan mute nasin en toki.
wawa tawa toki pi jan sona.
pi ilo nanpa en nanpa nasin.
taso tawa toki sona a.
sitelen sona, sitelen musi.

ilo lon Poki~\ref{hacks} li pana e sitelen
e sona tan sitelen,
en nasin.
Poki~\ref{basics} li pana e nasin pi toki pona.
e sitelen tawa kama sona.

mi wile pali e sitelen lon toki pona.
sitelen lon nasin, sitelen sona,
sitelen musi.
ante e toki pona
la ona li nasa.
taso nasa li pona mute.
li pona mute tawa lawa,
tawa kama sona, tawa sitelen e toki.
ante la toki pona o.
toki e toki pona tawa sina.
kama sona e toki pona sina.

o pona tawa jan pi toki pona.
tawa jan Sonja, Birns-Sprage, Kipo, Pije,
Siwejo, Malija, Tepan, jan kulupu mute.

>>>>

\section{Conclusions and further work}\label{conc}
This document presented a potentially novel description of Toki Pona (TP)
language in Section~\ref{basics}, and innovative and useful software routines for dealing with
the language in Section~\ref{hacks}.
As a minimalist conlang, TP is very convenient for cognitive
experiments
through linguistic semantics and
for devising tools for analysis, creation and visualization,
a fact to which this document is evidence.
Other conlangs, and even non-planned languages might benefit from
TP and this content in many ways, e.g. one might
synthesize TP texts and translate them as convenient,
adapt the routines to obtain statistics of the vocabulary,
or take advantage of the language description to devise new conlangs
or stylistic guidelines for any language.

Some of the many potential next steps should be highlighted:

@@

  \item The text synthesis facilities described in

@@

  \item Syntax highlighting that colors the tokens with respect
    to the syntactic positions and functions, and not in a fixed manner
    as is implemented and exposed in Section~\ref{shigh},
    may be implemented by using n-grams and
    other techniques from Natural Language Processing.
  \item The TP Wordnet(s) described in Section~\ref{wn} should be may
    be enhanced:
    implement a NLTK-like Wordnet interface; double-check if each synset

@@

    semicolon; scrutinize the neighborhood of the synsets of the TP
    Wordnet in the English Wordnet.

@@

    use insights derived from TP Wordnet.
    Create conlangs for specific uses:

@@

    A conlang might have different modes that are
    specified by section headers or tags. 
  \item Explore TP relation to other languages

@@

    and the reasons that led Sonja (and maybe other people)

@@

  \item Write academic articles in TP

@@

    facilitated acquisition of context,
    and an article in a canonical structure
    (e.g. introduction and related work, materials and methods,

@@

    Maybe something about complexity, statistics, physics,
    or computer science;
    or linguistics, philosophy, literature
    or psychology.
    Partners may learn the language in a few hours or weeks
    (with or without help).
    This is a very challenging step.

@@

  \item Explore a minimalistic conlang to describe software

@@

    In this sense, it seems advantageous

@@

    Consider the possibilities of and reasons for parsing TP.

@@

\appendix
\section{Current usage of Toki Pona by the author}\label{mytoki}
Disclaimer: this whole section holds personal views and usages of
the Toki Pona (TP) language.
It is reasonable to use the standard sounds, but often [z] instead of s
is more comfortable.
Translation into Toki Pona (e.g. of biblical excepts),
and the creation of new texts as poems and short stories,
is both challenging and pleasant.
Texts are shared e.g. in~\cite{tokisona}
and/or published in the TP Facebook groups.
The li particle is most often not used after subjects sina and mi,
in accordance with the norm,
but, even so, they are useful when there are many predicates.
E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
In such cases, the first li is sometimes omitted.
Also, li is found after mi and sina where one finds
that there is unwanted ambiguity, e.g.
sina moku pona 
which 
might be 'sina li moku pona' or 'sina moku pona (li jo e sike)'.

Names are by default transliterated,
but one should notice that, as in other languages,
names might be used as they are in the
correspondent mother tongue.
E.g. the name Erdös is used in
English and Portuguese although the standard
alphabet does not contain ö in such languages.
The use of e.g. English (or German) words
in TP texts is found when needed,
as happens often in scientific writing
(kernel is a German word used in English,
webpage and software are English words used in Portuguese).

Proposed notations for numbers seem numerous.
E.g. one might indicate if two numbers
are being multiplied (pi) or are in different scales
(such as in decimal or binary notations):
luka (pi) two mean 52.
Most importantly,
52 is a reasonable notation for a simple language.
E.g.
'mi jo e jan sama nanpa 12',
or
'ona li lon e soweli 27'.

Avoiding needless words...
'e ni:' may most often be written as ':'.
Subject phrase may be omitted (sometimes also the li)
if it is the
same as in the last sentence.
One may vary TP by grouping e.g. these words:
noka and anpa; luka, suno, sike and lawa; pali and pana.

Also, ambiguities introduced by omission of li
and absence of a token to denote preposition,
suggest the possibility of a syntax that is always uniquely
parsed (or at least less ambiguous).

It is tempting to conceive a language with the same syntax of TP:
'sentence' la 'sentence = subject + predicate + object' (each term with
a possible prepositional complement),
but always using particles to bound the sentence sections,
repeating them when the function is performed in the smallest scale:
'toki pi toki pona pi jan sona' meaning
language in toki pona and of intellectuals vs
'toki pi toki pona pi pi jan sona'
meaning language in [toki pona of intelectuals].
The keyword for initiating a preposition complement 
might be 'a', but it is already taken in TP.
One possibility is to use 'a' before a preposition and use 'ha' or 'he'
instead of TP a.
In such a setting,
one might enable pi, li, e, a (concept-qualifier, subject-predicate,
predicate-object, concept-preposition)
in any slot of the syntax template.
Between sentences one may use:
la pi, la li, la e, la a, and la la.
This is a very fractal conlang proposal.
Maybe also have a way to discern between a noun,
a qualifier (adjectives/adverbs) and a verb,
and accept any of them for the subject, predicate, object, preposition,
slots. Or assume a part-of-speech as default for a slot or for the
first word in a slot.

\section{Final words in Toki Pona}\label{ftp}
toki li nasin e lawa.\\
li nasin tawa (pi) toki insa.

% jan li ken toki li ken pali pona e toki pona.\\
toki pona li pona e nasin tan ni:\\
ona li jo e nimi nanpa lili.\\
ona li pona.\\
pona kepeken weka 'p' li ona, a.

o taso la toki pona li kalama li lukin pona.\\
sitelen en nasin li open e sitelen\\
suli~\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.\\
li sona. li nasin e toki e sona e lawa e lon.

toki ni li wawa tawa jan mute nasin en toki.\\
wawa tawa toki pi jan sona.\\
pi ilo nanpa en nanpa nasin.\\
taso tawa toki sona a.\\
sitelen sona, sitelen musi.

ilo lon Poki~\ref{hacks} li pana e sitelen\\
e sona tan sitelen,\\
en nasin.\\
Poki~\ref{basics} li pana e nasin pi toki pona.\\
e sitelen tawa kama sona.

mi wile pali e sitelen lon toki pona.\\
sitelen lon nasin, sitelen sona,\\
sitelen musi.\\
ante e toki pona\\
la ona li nasa.\\
taso nasa li pona mute.\\
li pona mute tawa lawa,\\
tawa kama sona, tawa sitelen e toki.\\
ante la toki pona o.\\
toki e toki pona tawa sina.\\
kama sona e toki pona sina.

o pona tawa jan pi toki pona.\\
tawa jan Sonja, Birns-Sprage, Kipo, Pije,\\
Siwejo, Malija, Tepan, jan kulupu mute.
